---
title: "Using functions, functionals, for loops, and dplyr to write concise code"
author: "Lisa Lipani"
date: "3/29/2019"
output: html_document
---

## Reading in files

* Reading in files individually is cumbersome and can lead to errors.

```{r, echo=TRUE}
at0H0_1_oldway <- read.delim("/users/lisalipani/Documents/School/Graduate/Research/NSP/output/RainbowAcoustics/at0H0_1_acoustics", col.names=c("voicelessFrames", "F0","harmonicsToNoise","zScoreRMS","start","end"), sep ="\t", header = FALSE, na.strings = "--undefined--")

at0H0_2_oldway <- read.delim("/users/lisalipani/Documents/School/Graduate/Research/NSP/output/RainbowAcoustics/at0H0_2_acoustics", col.names=c("voicelessFrames", "F0","harmonicsToNoise","zScoreRMS","start","end"), sep ="\t", header = FALSE, na.strings = "--undefined--")

# etc.
```

## Reading in files concisely

* A more concise way to read in multiple files is to use functionals.

### Make a vector

* First, we'll get the names of the filepaths with ```Sys.glob```. This creates a vector of the filepath names.

```{r}
# read in all file paths in folder to a list
files <- Sys.glob("/users/lisalipani/Documents/School/Graduate/Research/NSP/output/RainbowAcoustics/*")
head(files)
```

### Using ```lapply```
* We can use ```lapply``` to create a **list** of **data frames** for each file that we read in.
* First, let's learn about ```lapply```.
* ```lapply``` takes two arguments, a list or vector and a function, and it applies the function to the list or vector
    + A list can contain values of different types.
    + A vector only contains elements of one type.
  
### Writing functions
* Since ```lapply``` takes a function as its argument, we'll need to supply it with a function.
* When you write a function, you'll specify the arguments of your function in parenthesis, and what you want the function to do will go in curly brackets.
* Let's say you wanted to create a function that added two numbers together:

```{r}
addition <- function(x, y){
  return(x + y)
}

addition(2, 1)
```

* We've created a function called ```addition``` that takes arguments ```x``` and ```y``` and then does what you have specified in the curly brackets. 
* ```return``` isn't totally necessary here

### Using ```lapply``` and your function together

* ```lapply``` is a **functional**
    + A functional is a function that takes a function as an argument and returns a vector as output
    + ```sapply``` returns a vector, matrix, or array
    + ```vapply``` works the same way as ```sapply``` but you specify what you want R to return (a vector, matrix, or array)
* Here, ```lapply``` takes the vector named ```files``` and applies a function we wrote.
* The function we wrote takes one argument, ```i```

```{r}
# applies the function we define to everything in the list 'files'
all_files_list <- lapply(files, function(i){
  read.delim(i, col.names=c("voicelessFrames", "F0", "harmonicsToNoise", "zScoreRMS", "start", "end"), sep ="\t", header = FALSE, na.strings = "--undefined--")
})
```

* You could also name the function first, and then use it as an argument to ```lapply```
```{r}
# another way
# named function
function_to_read_files <- function(i){
read.delim(i, col.names=c("voicelessFrames", "F0", "harmonicsToNoise", "zScoreRMS", "start", "end"), sep ="\t", header = FALSE, na.strings = "--undefined--")
}

all_files_list_2 <- lapply(files, function_to_read_files)
```

* You could also pass the arguments of the function in ```lapply``` to ```lapply``` itself

```{r}
# another way
all_files_list_3 <- lapply(files, read.delim, col.names=c("voicelessFrames", "F0", "harmonicsToNoise", "zScoreRMS", "start", "end"), sep ="\t", header = FALSE, na.strings = "--undefined--")
```

```{r}
# check to see if output was the same
identical(all_files_list, all_files_list_2)
identical(all_files_list_2, all_files_list_3)
```
### Naming

* Once we have a list of data frames, it might be nice to name the elements

```{r}
library(tidyverse)
# use stringr package (a part of tidyverse) to find the last part of the file path (i.e. get the name of the file)
names_of_files <- str_match(files, "[^/]+$")

names(all_files_list) <- names_of_files
```

### Subsetting

* Once we have a list of data frames, we might need to access elements of it by subsetting. 

```{r}
# returns the head of the first data frame created
head(all_files_list[[1]])
```
```{r eval = FALSE}
# this won't do what I want it to
head(all_files_list[1])
```

### For loops

* We can also read files in with a for loop, where objects in the environment will be created

#### What is a for loop?
```{r}
# the basics of a for loop
for (i in 1:5){
  # do this thing five times
  print(i + 2)
}
```
```{r}
list_example <- c("a", "b", "c")

for (i in list_example){
  print (i)
}
```

### Create the loop and tell it what to repeat
```{r}
# sets up an iteration variable
i <- 0

# for each item (which is called 'name') in the list names, do the following things
for (name in names_of_files){
  # increases the iteration variable each time the loop is completed
  i <- i+1
  # assigns the complete filepath of the file to a variable named "filepath"
  filepath <- file.path(files[i])
  # assigns the results of read.delim() to the name of the file
  assign(name, read.delim(filepath, col.names=c("voicelessFrames", "F0", "harmonicsToNoise", "zScoreRMS", "start", "end"), sep ="\t", header = FALSE, na.strings = "--undefined--"))
}
```

## Replacing NAs in your data: Another example of ```lapply```

* Say you want to replace all the NAs in your data frames in the list we just created. We can use ```lapply``` and the function ```subsituteNA``` from the package ```timeSeries```.

```{r}
library(timeSeries)
# create a list of data frames that don't have NA
all_files_list_wo_NA <- lapply(all_files_list, substituteNA)

head(all_files_list[[1]])
head(all_files_list_wo_NA[[1]])
```

## Computing a dissimilarity matrix: Nested for loops

```{r}
# computes a dissimilarity matrix using dynamic time warping

# load in package dtw 
library(dtw)

# create a blank matrix to be populated
result1 <- matrix(, nrow = 60, ncol = 60)

# we created ```names_of_files``` earlier to give the data frames in a list names
# now, I'm creating "sentence1" to just get data frames containing data from a certain sentence
sentence1 <- all_files_list_wo_NA[grepl("_1_", names_of_files)]

# create your function
comp_diss_matrix <- function(sentence, result) {
  for (i in 1:length(sentence1)) {
    for (j in 1:i) {
      result[i, j] <- dtw(sentence[[i]][,c(1:5)], sentence[[j]][,c(1:5)], distance.only = TRUE)$normalizedDistance
    }
  }
  return(result)
}

# run the function
result1 <- comp_diss_matrix(sentence1, result1)

```
first loop  | second loop
------------|--------------
1           | 1
2           | 1, 2
3           | 1, 2, 3
4           | 1, 2, 3, 4
5           | 1, 2, 3, 4, 5

```{r}
# to make it faster
library(compiler)

compiled_diss_matrix <- cmpfun(comp_diss_matrix)

system.time(result1 <- compiled_diss_matrix(sentence1, result1))

system.time(result1 <- comp_diss_matrix(sentence1, result1))

head(result1)
heatmap(result1, Rowv = NA, Colv = "Rowv")
```

## Subsetting data

* Again, the old way is cumbersome and can lead to errors
```{r}
# old way
rainbow.df <- read.delim("/Users/lisalipani/Documents/School/Graduate/Research/NSP/Output/RainbowVoicingResults.txt", header = FALSE, col.names=c("sex", "file", "word", "phone", "phone_start", "phone_end", "word_start", "word_end", "duration", "previous_phone", "next_phone", "voicing_1", "voicing_2", "voicing_3", "voicing_total"))

rainbow_female_oldway <- subset(rainbow.df, rainbow.df$sex == "female")
rainbow_male_oldway <- subset(rainbow.df, rainbow.df$sex == "male")
# etc.
```

## Subsetting data using dplyr

```{r}
rainbow_female <- rainbow.df %>% 
  # could also use just 'filter' here, but this will work if you have a package loaded that masks filter
  dplyr::filter(sex == "female")

head(rainbow_female)
```
* There's lots of ways you can subset your data and return responses
* And there's lots of good information on how to use ```dplyr``` on this [cheatsheet](https://ugoproto.github.io/ugo_r_doc/dplyr.pdf) and [tidyverse](https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf) in general.
```{r}
# get the mean duration for the phone [t] grouped by sex
rainbow.df %>%
  group_by(sex) %>%
  dplyr::filter(phone == "t") %>%
  summarise(avg = mean(duration))
```



